A practical look at genetic algorithms

### 2015-12-08 evolutionaryComputation/1-practical/report.tex
split up the practical in 2 folders

\paragraph{Fittest filter} The fittest filter decides how which of both
the parents and children continues.
The most simple implementation is the \emph{Killparents}, which
returns the children. \emph{Truncate elitism} selects the best of both parents
and children. \emph{Tournament elitism} matches child next to each parent
and decide which one is best. There is no kill children implementation
because nothing would ever change.

@@

\paragraph{Member factories} The member factories are able to create
members from strings. These are either the tightly linked or randomly
linked. 
\subparagraph{Mutation} Then the tightly or randomly linked
functions can also be wrapped in a mutation function,
which will mutate the string before passing it to the factory.
Its an decorator idea, only with functions.

@@

The experiment is setup to do the bisectional search. The bisectional search
function will wrap the result of \emph{findOptimum} in a \emph{StochasticRun}.
\emph{findOptimum} is a recursive function that calls verify lowest to get a
collection of runResults. This is used to determine if the population size
should be increased or decreased like a bisection search.
\emph{verifyLowest} will try to run a population size 30 times, if the faults
limit is exceeded or the index exceeds the required runs limit it will quit.
In the latter case the run is considered to be successful.
The run will execute the GA search according to the configuration of the 
experiment, so this function will call startGenetic, and put the results
of the run into a RunResult. The RunResult will actually truncate a lot of
information. Not the entire resulting population is stored. It wasn't
necessary for the experiment and I was concerned about memory. Or rather
I didn't want to be too wasteful, especially considering that the JVM
reserves big blocks of memory.

\subsubsection{Unit tests}
Because the assignment has very well defined behavior. I thought it was
a very wise Idea to use unit tests. To do this I use a library called
Scala test.

I found about 6 different bugs alone in the valuation functions by using 
unit-testing, so I think this saved me considerable time.
\subsubsection{Build tool}
I use Gradle for building the project. Its the most advanced build platform
available and makes it easy to add dependencies to the project.
For example I use slf4j for logging.

@@

\subsubsection{Libraries used}
In here I will describe why I chose to use these libraries
The details of the libraries can be found in the gradle file (which
version and what maven group):

@@

\subsection{Speed up attempts}
I wanted my implementation to be fast, preferably faster than any other
implementation, now I think I didn't succeed in this because I used a
lot of slow mechanisms, such as functional combinators, and although
they don't increase the computational complexity, they do certainly increase
the minimum required runtime, I'm quite sure that every combinator in sequence
is implemented as just another loop. However because I
run all experiments in parallel the total time may be faster.
\subsubsection{Parallelization}
Since were doing string comparisons I had the idea of using parallelization to
speed up the algorithm by doing parallelization. This can be done very easily
in Scala with par collections (and the fact that immutability is default).

However once I implemented this, I came to the conclusion it was not faster.
The overhead for delegating and creating threads was more than the parallelization
gained.  After 10 minutes of waiting I decided to drop this idea
since without parallelization it took about 2 minutes (I tried this
on just one run, so it shouldn't take that long).

The problem is that the evaluation function, even with gene length of 100, is
to cheap to make any reasonable parallelization effective. However what was effictive
is the parallelization of the experiments. This reduced total runtime
significantly. However this is not seen in the statistics.

So now all experiments are run at the same time. This probably increases the
average time of each `run' somewhat (not that much because the concurrent
collection does \href{http://docs.scala-lang.org/overviews/parallel-collections/configuration.html}{load balancing}).
It also drastically decreases the amount of time I have to wait. Which is quite
important, especially in the development phase. (although I also used
other tricks, such as instead of 30, just doing 3 runs)

@@

\subsubsection{Only create children if the genes are new}
If the genes are new then create the children otherwise return the parent.
This reduces evaluations and object creations. Its becomes more effective
at deeper generation, it even slows down the fastest GA's by a few milliseconds
but the deeper GA are lots faster.

@@

To reduce the amount of string comparisons, first the hash code function is called
This will cache itself, I know this because I inspected the source code of the
String class in Java.

I was very proud for coming up with  this idea, but when I ran the experiment with
and without hash code I came to the conclusion that first doing a hash code operation
significantly *slows* execution. This becomes worse as population size increases.

\section{Experiment}
Something to note is that my made up tournament elitism is actually in most
cases better, or at least equal to the truncate elitism. So when I saw
this I tried using my other initially used algorithm where I would completely
ignore the parents (kill the parents). However it appears this does not
perform better.

I also added two point cross with mutation and 1 point cross to see how
those fair, running the code now takes about 40 minutes in total with the
entire thing divided over 8 threads (quad core + hyper-threading). The only
other improvement I could think of is running it on the GPU, but I think
the algorithm is to complex for that and I don't know how to do that,
especially considering I'm using the JVM.

Two point cross and single cross both seem to find results in deceptive
random tournament elitism.
\subsection{Hypothesis}
The deceptive trap function will not be solved by any algorithm for
the population sizes. I think this because its to desirable to have
at least one 0000 block in the 100 string length.
Non-deceptive will be more successful. The others will be solved.
I also think mutation will have an important effect and may even
allow deceptive to be solved.

I also think that the provided selection pressure would be about
average. It will probably always find a result.
\subsection{Execution}
I added extra selection methods and crossing strategies to
the final experiment. I did this because I was curious how they would
compete with the original assignment.

@@

\paragraph{Abreviations}
\begin{tabular}{ll}
	Abrivation & Full name \\ \toprule
	2X & 2 Point cross \\
	2XB & 2 Point cross biased \\
	2XM & 2 Point cross mutated \\
	2XM & 2 Point cross mutated \\
	UX & uniform cross \\
	UXM & uniform cross mutated \\
	1X & 1 Point cross \\ \bottomrule
\end{tabular}

@@

It turns out however that the biased uniform cross is much more effective.
As it solves the deciptive trap function and leaves the `normal' one in
the dust. I actually excpected an improvement when changing the biased
variant into the current vairant. To my shock it did major damage to the
algoritm. To show this I decided to include both.

\subsubsection{Graph observations}
The major drops are caused by an optimization I implemented. After 2 failed
run the stocastic ensurance function will stop. Often this happens when
zooming in, this makes it appear as if there is a local gap, but its just
random chanche and the fact that the bisectional search is zooming in.

@@

\section{Conclusion}
%What did I learn from my observation?
My conclusion is that using both tournament selection and truncate elitism
is to harsh. It will slow down the GA and often not find a solution at all
(in case of deceptive trap, 2x should've come up with something). the 
time vs population found ratio is also much better for just using tournament
select.

Whats also interesting is that if you just throw away the parents and use 
the children (kill parents), then you still will often perform better than
using tournament selection and truncate elitism, although not as 
consistently as tournament elitism.

To my surprise deceptive trap was solved with decreased selection pressure
by both 2x and 1x, although 1x did worse. Also adding mutation doesn't do
the algorithm much good. 2XM and UXM consistently find worse minimum populations
than respectively 2X and UX{.}
This probably means the mutation is too disruptive.

Mutation turned out to have no significant effect other than just slowing
things down.

>>>>

\label{paragraph:Fittest filter}
\paragraph{Fittest filter} The fittest filter decides how which of both
the parents and children continues.
The most simple implementation is the \emph{Killparents}, which
returns the children. \emph{Truncate elitism} selects the best of both parents
and children. \emph{Tournament elitism} matches child next to each parent
and decide which one is best. There is no kill children implementation
because nothing would ever change.

@@

\paragraph{Member factories} The member factories are able to create
members from strings. These are either the tightly linked or randomly
linked.

\subparagraph{Mutation} Then the tightly or randomly linked
functions can also be wrapped in a mutation function,
which will mutate the string before passing it to the factory.
Its an decorator idea, only with functions.
\newpage

\subsubsection{Experiment}

@@

The experiment is setup to do the bisectional search. The bisectional search
function will wrap the result of \emph{findOptimum} in a \emph{StochasticRun}.
\emph{findOptimum} is a recursive function that calls verify lowest to get a
collection of runResults. This is used to determine if the population size
should be increased or decreased like a bisection search.
\emph{verifyLowest} will try to run a population size 30 times, if the faults
limit is exceeded or the index exceeds the required run limit it will quit.
The run is considered to be successful if the count of successful run >= 
run limit - faults.

The run will execute the \emph{Evolution} that has been passed into this
experiment.  so this function will call startGenetic on the Evolution object, 
and put the results of the run into a \emph{RunResult}.

\subsection{Unit tests}
Because the assignment has very well defined behavior. I thought it was
a good Idea to use unit tests. To do this I use a library called
Scala test.

I found about 6 different bugs alone in the valuation functions by using
unit-testing, so I think this saved me considerable time. I also used this
to precisely define the behavior of 2 point cross.

By using unit tests I became very certain about the execution of my code.
It does precisely what its supposed to do even though I use functional
programming and it may look somewhat foreign.

\subsection{Build tool}
I use Gradle for building the project. Its the most advanced build platform
available and makes it easy to add dependencies to the project.
For example I use slf4j for logging.

@@

\subsection{Libraries used}
In here I will describe why I chose to use these libraries
The details of the libraries can be found in the Gradle file (which
version and what maven group):

@@

\subsection{Speed up attempts}
I wanted my implementation to be fast, preferably faster than any other
implementation. I think I didn't succeed in this for the algorithm part
because I used a lot of slow mechanisms, such as functional combinators.
Although they don't increase the computational complexity,
they do certainly increase
the minimum required runtime. I'm quite sure that every combinator in sequence
is implemented as just another loop. However because I
run all experiments in parallel the total time may be faster. So in a way
I did succeed.

\subsubsection{Parallelization}
Since were doing string comparisons I had the idea of using parallelization to
speed up the algorithm by doing parallelization. This can be done very easily
in Scala with par collections (and the fact that immutability is default).

However once I implemented this, I came to the conclusion it was not faster.
The overhead for delegating and creating threads was more than the parallelization
gained.  After 10 minutes of waiting I decided to drop this idea
since without parallelization it took about 2 minutes.
The problem is that the evaluation function, even with gene length of 100, is
to cheap to make any reasonable parallelization effective.

So now all experiments are run at the same time. This probably increases the
average time of each `run' somewhat (not that much because the concurrent
collection does \href{http://docs.scala-lang.org/overviews/parallel-collections/configuration.html}{load balancing}).
It also drastically decreases the amount of time I have to wait. Which is quite
important, especially in the development phase.

@@

\subsubsection{Only create children if the genes are new}
This optimization reduces the amount of evaluations required.  It becomes 
more effective at deeper generations, however it slows down the fastest 
GA's by a few milliseconds but the slower GA's  are quite a bit faster.

@@

I also had an idea of using hash-codes to reduce the amount of comparisons.
But it turns out that this increases the running time, so I quickly removed
this optimization again.

\section{Experiment}
In here I will discuss how the experiments ran and the thoughts I
above about the results.
First the setup will be discussed, since I added to the
original assignment out of curiosity. Then the hypothesis
will be discussed. After that the results will be shown
and finally I will discuss the conclusions I draw from the experiments.

\subsection{Experiment titles}
The experiment titles all follow the same structure:

\[F\mbox{ }L, E\]
Where F is the fitness function, L is the linkage and E is the elitism type.
The experiment results are grouped by this structure alphabetically.
The elitism types are discussed in the fittest fileter paragraph \ref{paragraph:Fittest filter}

Block is the trap function.

\subsection{Abbreviations}
\begin{tabular}{ll}
	Abrivation & Full name \\ \toprule
	2X & 2 Point cross \\
	2XB & 2 Point cross biased \\
	2XM & 2 Point cross mutated \\
	2XM & 2 Point cross mutated \\
	UX & uniform cross \\
	UXM & uniform cross mutated \\
	1X & 1 Point cross \\ \bottomrule
\end{tabular}

\subsection{Setup}
I added extra selection methods and crossing strategies to
the final experiment. I did this because I was curious how they would
compete with the original assignment.

@@

@@

It turns out however that the biased uniform cross is much more effective.
As it solves the deceptive trap function and leaves the `normal' one in
the dust. I actually expected an improvement when changing the biased
variant into the current variant. It was significantly worse however.
To show this I decided to include both.

1 point cross was added to see how that competes with 2 point cross.
I also used to have 2 point cross with
mutation but it wasn't significantly different then the normal 2X
and the graphs started to get cluttered so I decided to drop that in
favour of discussing the biased crossover.

Running the code now takes about 26 minutes in total with the
entire thing divided over 8 threads (quad core + hyper-threading).

\subsection{Hypothesis}
The deceptive trap function will not be solved by any algorithm for
the population sizes. I think this because its to desirable to have
at least one 0000 block in the 100 string length.
Non-deceptive will be more successful. The others will definitively be solved.
I also think mutation will have an important effect and may even
allow deceptive to be solved.

The provided selection pressure would be about average.
It will probably always find a result.

@@

\subsubsection{Graph observations}
The major drops are caused by an optimization I implemented. After 2 failed
run the stochastic insurance function will stop.

Often this happens when the bisectional search hones in on the optimum
This makes it appear as if there is a local gap, but its just
random chance.

\section{Conclusion}
My conclusion is that using both tournament selection and truncate elitism
is to harsh. It will slow down the GA and often not find a solution at all.
This happens for example with the deceptive trap function, randomly and
tightly linked.

The time vs population found ratio is also much better for using 
tournament select on both the elitism and selection.

Whats also interesting is that if you just throw away the parents and use
the children (kill parents), then you still will often perform better than
using tournament selection and truncate elitism (this is how the original
assignment was), although not as consistently as tournament elitism.

To my surprise deceptive trap was solved with decreased selection pressure
by both 2x and 1x, although 1x did worse.

Mutation doesn't seem to have much effect UXM sometimes finds a better
solution but usually this is not the case. I initially thought mutation
would be very important but its just sort off random.  Mutation turned out
to have no significant effect other than just slowing things down.

Something to note is that my made up tournament elitism is actually in most
cases better, or at least equal to the truncate elitism. So when I saw
this I tried using my other initially used algorithm where I would completely
ignore the parents (kill the parents). However it appears this does not
perform better.

Two point cross and single cross both seem to find results in deceptive
random tournament elitism.

Linearly scaled and uniformly scaled are very easy to solve for almost all
algorithms. Although the 2x and 1x are at a big disadvantage compared to UX.
